---
title: "Global $CO_{2}$ Emissions in 1997"
short: "What Keeling missed all these years"
journal: "AER" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
vol: 0
issue: 0
keywords:
  - Global Warming
  - Keeling Curve
author:
  - name: Meral Basit
    firstname: Meral
    surname: Basit
    email: meral@ischool.berkeley.edu
    affiliation: UC Berkeley, School of Information
  - name: Alex Hubbard
    firstname: Alex
    surname: Hubbard
    email: alex.hubbard@ischool.berkeley.edu
    affiliation: UC Berkeley, School of Information
  - name: Mohamed Bakr
    firstname: Mohamed 
    surname: Bakr
    email: m.bakr@ischool.berkeley.edu
    affiliation: UC Berkeley, School of Information
acknowledgements: | 
  Acknowledgements
abstract: | 
  Abstract goes here
header-includes: 
  - '\usepackage{graphicx}'
  - '\usepackage{booktabs}'
  # - '\usepackage[numbers]{natbib}'
output:  pdf_document
# output: rticles::aea_article
bibliography: BibFile.bib
# biblio-style: apalike
csl: data-science-journal.csl
---

```{r setup, echo=FALSE}
## default to not show code, unless we ask for it.
knitr::opts_chunk$set(echo=FALSE)
options(digits = 3)
```

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
```

\newpage

# Background

## Carbon Emissions

Carbon emissions refer to the release of carbon, particularly carbon dioxide ($CO_2$), into the atmosphere. This process primarily occurs through the burning of fossil fuels such as coal, oil, and natural gas, as well as through deforestation and various industrial processes. $CO_2$ is a greenhouse gas, meaning it traps heat in the Earth's atmosphere and contributes to the greenhouse effect, which leads to global warming and climate change.

In our report we are trying to understand the trend of the atmospheric CO2 by asking the following research question:

```{=tex}
\begin{quote}
  \textit{Is there a significant upward trend in atmospheric CO2 levels over time?}
\end{quote}
```
### Null Hypothesis

There is no significant upward trend in atmospheric CO2 levels over time.
$H_0: \beta_1 \leq 0$
Where: 
$\beta_1$ is the trend coefficient over time in a linear regression model of the form $CO2_t = \beta_0 + \beta_1 \cdot t + \epsilon_t$.
$CO2_t$ is the atmospheric $CO_2$ level at time $t$.

# Measurement and Data

## Measuring Atmospheric Carbon

In this study, we will use the Mauna Loa Atmospheric $CO_2$ Concentration time series dataset that is a  available in `R`. The time series of 468 observations of the monthly Atmospheric concentrations of $CO_2$  from 1959 to 1997 expressed in parts per million (ppm). This means that, for example, a value of 320 means there are 320 CO2 molecules for every 1 million air particles (after the water vapor is removed from the sample). The data was reported in the preliminary 1997 SIO [@keeling1997].

The data where collected at the Mouna Loa Observatory which located on the island of Hawaii at an elevation of 11,135 feet above sea level which makes his location is "well situated to measure air masses that are representative of very large areas" [@4v5dl8o8qp].

The values for February, March and April of 1964 were missing and have been obtained by interpolating linearly between the values for January and May of 1964.

## Historical Trends in Atmospheric Carbon


# Models and Forecasts

## Linear Models

## ARIMA Models

## Forecasts

# Conclusions

# Refrences

\bibliographystyle{aea}
\bibliography{BibFile}
