---
title: "Global $CO_{2}$ Emissions in 1997"
short: "What Keeling missed all these years"
author: "Meral Basit, Alex Hubbard, Mohamed Bakr"
output:
  bookdown::pdf_document2:
    toc: true
    # citation_package: natbib
    fig_caption: true
    latex_engine: pdflatex
fontsize: 11pt
geometry: margin=1in
editor_options: 
  chunk_output_type: inline
bibliography: BibFile.bib
csl: data-science-journal.csl
---

```{r setup, echo=FALSE}
## default to not show code, unless we ask for it.
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE, dpi=1000)
options(digits = 3)
```

```{r load packages, echo = FALSE, message = FALSE}
library(tidyverse)
library(tsibble)
library(latex2exp)
if (!requireNamespace("forecast", quietly = TRUE)) {
  install.packages("forecast")
}
library(forecast)
library(patchwork)
library(fable)
library(lubridate)
library(fable )
library(feasts)
library(astsa)
library(patchwork)
library(car)

theme_set(theme_minimal())
```

\newpage

# Background

## Carbon Emissions

Carbon emissions refer to the release of carbon, particularly carbon dioxide ($CO_2$), into the atmosphere. This process primarily occurs through the burning of fossil fuels such as coal, oil, and natural gas, as well as through deforestation and various industrial processes. $CO_2$ is a greenhouse gas, meaning it traps heat in the Earth's atmosphere and contributes to the greenhouse effect, which leads to global warming and climate change.

In our report we are trying to understand the trend of the atmospheric CO2 by asking the following research question:

```{=tex}
\begin{quote}
  \textit{Is there a significant upward trend in atmospheric CO2 levels over time?}
\end{quote}
```
### Null Hypothesis

There is no significant upward trend in atmospheric CO2 levels over time. $H_0: \beta_1 \leq 0$ Where: $\beta_1$ is the trend coefficient over time in a linear regression model of the form $CO2_t = \beta_0 + \beta_1 \cdot t + \epsilon_t$. $CO2_t$ is the atmospheric $CO_2$ level at time $t$.

# Measurement and Data

<!-- (3 points) Task 1a: CO2 data -->
<!-- Conduct a comprehensive Exploratory Data Analysis on the co2 series. This should include (without being limited to) a description of how, where and why the data is generated, a thorough investigation of the trend, seasonal and irregular elements. Trends both in levels and growth rates should be discussed (consider expressing longer-run growth rates as annualized averages). -->
<!-- What you report in the deliverable should not be your own process of discovery, but rather a guided discussion that you have constructed so that your audience can come to an understanding as succinctly and successfully as possible. This means that figures should be thoughtfully constructed and what you learn from them should be discussed in text; to the extent that there is any raw output from your analysis, you should intend for people to read and interpret it, and you should write your own interpretation as well. -->

## Measuring Atmospheric Carbon

In this study, we will use the Mauna Loa Atmospheric $CO_2$ Concentration time series dataset that is a available in `R`. The time series of 468 observations of the monthly Atmospheric concentrations of $CO_2$ from 1959 to 1997 expressed in parts per million (ppm). This means that, for example, a value of 320 means there are 320 CO2 molecules for every 1 million air particles (after the water vapor is removed from the sample). The data was reported in the preliminary 1997 SIO [@keeling1997].

The data where collected at the Mouna Loa Observatory which located on the island of Hawaii at an elevation of 11,135 feet above sea level which makes his location is "well situated to measure air masses that are representative of very large areas" [@gml, @co2_measurements].

The values for February, March and April of 1964 were missing and have been obtained by interpolating linearly between the values for January and May of 1964.

## Historical Trends in Atmospheric Carbon

The Keeling Curve is a graph of the accumulation of carbon dioxide in the Earth's atmosphere based on continuous measurements taken at the Mauna Loa Observatory on the island of Hawaii from 1958 to the present day. The curve is named for the scientist Charles David Keeling, who started the monitoring program and supervised it until his death in 2005 [@noauthor_keeling_2024]

```{r EDA standard plots}
# turn into tsibble
co2.ts <- as_tsibble(co2)

# time series
p1 <- ggplot(co2.ts, aes(x = index, y = value)) +
  geom_line(linewidth = .5) +
  labs(title = "Atmospheric CO2 1959-1997", x = "Month and Year", y = "CO2 (ppm)") +
  theme_minimal()

# acf
p2 <- ggAcf(co2.ts$value) + ggtitle("ACF CO2")

# pacf
p3 <- ggPacf(co2.ts$value) + ggtitle("PACF CO2")

# histogram
p4 <- ggplot(co2.ts, aes(x = value)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Atmospheric CO2", x = "CO2 (ppm)", y = "Frequency")

# annual average
p5 <- as_tibble(co2.ts) %>%
  mutate(year = year(index)) %>%
  group_by(year) %>%
  summarise(avg_value = mean(value, na.rm=TRUE)) %>%
  ggplot(aes(x=year, y=avg_value)) +
  geom_point() +
  labs(title = "Average Yearly CO2 Levels", x = "Year", y = "Average CO2 (ppm)") +
  theme_minimal()

# make layout
pw <- (p1 + p4) /
  (p2 + p3) /
  p5
```

```{r EDA-plots, include=TRUE,fig.cap="Atmospheric CO2 EDA standard plots", fig.width=9, fig.height=5}
pw + plot_annotation(
  title = 'Atmospheric CO2 EDA plots',
  subtitle = 'Upward trend with clear seasonal pattern',
  caption = 'Source: Mauna Loa Atmospheric CO2 Concentration'
)
```

As we see in Figure \@ref(fig:EDA-plots) the time series in the top left, there is a pretty strong seasonality as well as a linear upward trend. This dataset is not mean stationary but it may be variance stationary.

The ACF tails off very slowly while the PACF drops shortly after lag 1 but still have oscillating pattern. This suggests that this series may have a unit root.

The histogram in Figure \@ref(fig:EDA-plots) top right shows that there is a wide range of values with a slight right skew. The yearly average plot on the bottom illustrates the linear trend in the data series more explicitly.

```{r EDA seasonality}
p6 <- co2.ts %>%
  ggplot(aes(x = index, y = value)) +
  geom_line(linewidth = .5) +
  labs(title = "Atmospheric CO2", x = "", y = "CO2 (ppm)") +
  theme_minimal()

p7 <- ggplot(subset(co2.ts), 
       aes(x = index, y = difference(value,6))) +
  geom_line(linewidth = .5) +
  labs(title = "Seasonally Differenced, Lag 6", x = "Date", y = "CO2 (ppm)") +
  theme_minimal()
```

```{r def-plot, include=TRUE, fig.cap="Seasonally Differenced CO2 Concentrations (Lag 6)", fig.width=9, fig.height=2}
p7
```

```{r stationarity test KPSS, eval=FALSE, include=FALSE}
co2.ts %>% 
  mutate(log_value = log(value)) %>%
  features(log_value, unitroot_kpss)

co2.ts %>% 
  mutate(diff_value = difference(value, 6)) %>%
  features(diff_value, unitroot_kpss)
```

Based on the Figure \@ref(fig:EDA-plots) and after examining the time series more closely, we noted that the seasonal cycle seemed to be 6 lags long, we applied a seasonal differencing of 6 lags.  This transformation resulted in a stationary series.

To confirm this we performed KPSS unit root test. With p-value of 0.1, fail to reject null hypothesis of stationarity, indicating that the transformed series is now stationary Figure \@ref(fig:def-plot).

# Models and Forecasts

In this section, we will analyze and compare two different models to gain a better understanding of the complex dynamics of the time series process. We will assess a linear model and an ARIMA model to identify the most appropriate time series model for our analysis.

## Linear Models

<!-- (3 points) Task 2a: Linear time trend model -->
<!-- Fit a linear time trend model to the co2 series, and examine the characteristics of the residuals. Compare this to a quadratic time trend model. Discuss whether a logarithmic transformation of the data would be appropriate. Fit a polynomial time trend model that incorporates seasonal dummy variables, and use this model to generate forecasts to the year 2020. -->

We start by fitting a linear model of the form: $$CO2_t=\phi_0+\phi_1t+\epsilon_t$$
Based on the fit results, the estimated coefficient $\beta_1 = 0.109$ indicates that the $CO_2$ levels increase by $\approx$ 0.109 units per month. The p-value of the time index is $<0.05$ which suggests that the coefficient is statistically significant. We reject the null hypothesis that the coefficient $\beta_1 = 0$. that also proves evidence that the $CO_2$ levels have an upward linear trend.

Both $R^2$ and the adjusted $R^2$ value is $0.969$, which means that the linear model can explain $96.9%$ of the $CO_2$ levels variance, suggesting that the model effectively captures the main patterns in the data.

```{r linear model}
mod.linear <- lm(value ~ time(index), data = co2.ts)
summary(mod.linear)

# Plot the linear model
lm.mod.plot <- co2.ts |> 
  ggplot(aes(x = index, y = value)) +
  geom_line() +
  geom_line(aes(y = predict(mod.linear)), color = "cornflowerblue") +
  labs(title = "CO2 Levels with Fitted Linear Time Trend Model", x = "Month", y = "CO2 (ppm)")

# Plot the residuals
lm.mod.res <- co2.ts |> 
ggplot( aes(x = index, y = resid(mod.linear))) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(formula = 'y ~ x',method = "loess", color = "blue", se = FALSE) +
  labs(title = "Linear Time Trend Model Residuals", x = "Month", y = "Residuals")

# residualPlot(mod.linear, main="Linear Residuals")
confint(mod.linear)
```

Next, we fit a quadratic model of the form: $$CO2_t= \phi_0 + \phi_1 t + \phi_2 t^2 + \epsilon_t$$
Based on the fit results of the quadratic time trend model, the estimated coefficient $\beta_1 = 0.0674$ indicates that the $CO_2$ levels increase by $\approx0.0674$ units per month. The p-value of the time index is $<0.05$ which suggests that the coefficient is statistically significant. We reject the null hypothesis that the coefficient $\beta_1 = 0$. that also proves evidence that the $CO_2$ levels have an upward linear trend.

the estimated quadratic term coefficient $\beta_2 = 0.0000886$ The positive coefficient suggests that the rate of increase in CO2 levels is accelerating. The p-value of the time index is $<0.05$ which suggests that the coefficient is statistically significant. We reject the null hypothesis that the coefficient $\beta_2 = 0$.

Both $R^2$ and the adjusted $R^2$ value is $0.979$, which means that the linear model can explain $97.9%$ of the $CO_2$ levels variance, suggesting that the model effectively captures the main patterns in the data.

After analyzing the two models, it is obvious that the significant coefficients show a clear upward trend in CO2 levels over time, with a rapidly increasing rate. However, it seems that the quadratic model is slightly outperforming the linear model when comparing the $R^2$ results and examining the residual plots in Figure \@ref(fig:res-compare).

```{r quadratic model}
mod.quadratic <- lm(value ~ time(index) + I(time(index)^2), data = co2.ts)
summary(mod.quadratic)

# Plot the quadratic model
qad.mod.plot <- co2.ts |>
ggplot(aes(x = index, y = value)) +
  geom_line() +
  geom_line(aes(y = predict(mod.quadratic)), color = "gold3") +
  labs(title = "CO2 Levels with Fitted Quadratic Time Trend Model", x = "Month", y = "CO2 (ppm)")

# Plot quadratic model residuals
qad.mod.res <- co2.ts |>
ggplot(aes(x = index, y = resid(mod.quadratic))) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(formula = 'y ~ x', method = "loess", color = "blue", se = FALSE) +
  labs(title = "Quadratic Time Trend Model Residuals", x = "Month", y = "Residuals")

# residualPlot(mod.quadratic, main="Quadratic residuals")
confint(mod.quadratic)
```
```{r res-compare, include=TRUE,fig.cap="Residuals of the Linear and Quadratic Models", fig.width=9, fig.height=3}
lm.mod.res + qad.mod.res
```

## ARIMA Models

## Forecasts

# Conclusions

# Refrences

```{=tex}
\bibliographystyle{aea}
\bibliography{BibFile}
```
